Inference of proto-neutron star properties from gravitational-wave data in core-collapse supernovae

### 2020-11-09 introduction.tex
update plots & text

The life of sufficiently massive stars, i.e.~those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$, ends with the collapse of {the} iron core under {its} own gravity, leading {to} the formation of a neutron star {(NS)} or a black hole (BH), {and} followed (typically but not necessarily in the BH case) by {a supernova} explosion. Nearby core-collapse supernova (CCSN) explosions are expected to be sources of gravitational waves (GWs) and they could be 
%are one of the main candidates for 
the next great discovery of current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion \citep{Bethe:1990} is the most likely outcome in the case of slowly rotating cores, which are present in the bulk of CCSN progenitors. The emitted GWs could be detected with the advanced ground-based GW detector network, Advanced LIGO (aLIGO)~\citep{TheLIGOScientific:2014jea}, Advanced Virgo (AdV)~\citep{TheVirgo:2014hva} and
KAGRA~\citep{Aso:2013eba}, within $\sim 5$~kpc \citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century \citep{Adams:2013,Rozwadowska:2021}.
For the case of rapidly rotating progenitor cores the result is likely a magneto-rotational explosion, yielding  a more powerful GW signal that could be detected within $50$~kpc and, for some extreme models, up to $5-30$ Mpc \citep{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe \citep{Li:2011b} or events associated with long GRBs 
\citep{Chapman:2007}) making this possibility a subdominant channel of detection with an event rate of $\sim 10^{-4} \rm{yr}^{-1}$\mab{[add ref?]}. For the results discussed in this work we only consider neutrino-driven CCSN. \tf{This last sentence must be mentioned somewhere (or maybe not) but perhaps here is not the best place.}\mab{MAB:I'm in favor of deleting it as it is rather obvious we are considering neutrino driven CCSN} Despite the low rates, CCSN are of great scientific interest because they produce complex GW signals which could provide significant clues about the physical processes at work after the gravitational collapse of stellar cores.

>>>>

The life of sufficiently massive stars, i.e.~those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$, ends with the collapse of {the} iron core under {its} own gravity, leading {to} the formation of a neutron star {(NS)} or a black hole (BH), {and} followed (typically but not necessarily in the BH case) by {a supernova} explosion. Nearby core-collapse supernova (CCSN) explosions are expected to be sources of gravitational waves (GWs) and they could be 
%are one of the main candidates for 
the next great discovery of current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion \citep{Bethe:1990} is the most likely outcome in the case of slowly rotating cores, which are present in the bulk of CCSN progenitors. The emitted GWs could be detected with the advanced ground-based GW detector network, Advanced LIGO (aLIGO)~\citep{TheLIGOScientific:2014jea}, Advanced Virgo (AdV)~\citep{TheVirgo:2014hva} and
KAGRA~\citep{Aso:2013eba}, within $\sim 5$~kpc \citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century \citep{Adams:2013,Rozwadowska:2021}.
For the case of rapidly rotating progenitor cores the result is likely a magneto-rotational explosion, yielding  a more powerful GW signal that could be detected within $50$~kpc and, for some extreme models, up to $5-30$ Mpc \citep{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe \citep{Li:2011b} or events associated with long GRBs 
\citep{Chapman:2007}) making this possibility a subdominant channel of detection with an event rate of $\sim 10^{-4} \rm{yr}^{-1}$\mab{[add ref?]}. For the results discussed in this work we only consider neutrino-driven CCSN.
%\tf{This last sentence must be mentioned somewhere (or maybe not) but perhaps here is not the best place.}\mab{MAB:I'm in favor of deleting it as it is rather obvious we are considering neutrino driven CCSN} Despite the low rates, CCSN are of great scientific interest because they produce complex GW signals which could provide significant clues about the physical processes at work after the gravitational collapse of stellar cores.

### 2020-11-09 results.tex
update plots & text

\begin{figure}[t]
 \centering
 \includegraphics[width=0.5\textwidth]{plots/spectrum}
 \caption{Amplitude spectral density of the GW detectors aLIGO and AdV at design sensitivity and that of the proposed third-generation detectors Cosmic Explorer and Einstein Telescope. Einstein Telescope sensitivity curve ET-B is obtained pushing second-generation detector technology at its limit. ET-C and ET-D sensitivity curves correspond to a detector configuration where a low-power cryogenic low-frequency interferometer and a high-power room temperature high-frequency interferometer are sharing the same infrastructure~\cite{Hild_2011}. Cosmic Explorer design sensitivity will be achieved in two stages. Stage 1 (CE1) is expected to use the technology developed for the ``A+'' upgrade to aLIGO but scaled up to a 40 km detector while stage 2 (CE2) will implement state-of-the-art technology to decrease quantum and thermal noises~\cite{reitze2019cosmic}. \tf{Use AdV in the legend instead of ADV.}} 
 \label{fig:spectrum}
\end{figure}

@@

To estimate how accurately we can infer the time evolution of $r(t)$ in the GW data of a single
detector we inject the GW signal for model {\texttt s20S} into 
100 Gaussian noise realisations whose power spectral density (PSD) follows the aLIGO
spectrum~\cite{aLIGOsens:2018}. The PSD of aLIGO and AdV at design sensitivities, along with those of planned third-generation detectors, are shown in Figure~\ref{fig:spectrum}.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/aLIGO_coverage_allwvfs}
 \caption{Median of the coverage for the eight CCSN waveforms of the {\it test set} embedded in aLIGO noise as a function of the  distance to the source. \mab{The ``no signal'' line shows the median of coverage in absence of any signal. In this example, the median is null and overlaps with the horizontal axis. The blue band boundaries are given by the $\mathrm{5^{th}}$ and $\mathrm{95^{th}}$ percentiles of coverage in absence of any signal.} \tf{Perhaps is more consistent to use in the legend the same model names as in Table 1. Some of the colours are really hard to see, particularly for s20S. That could perhaps be improved if we use thicker lines. I'm confused about the quartiles shown.}} \label{fig:aLIGO_cov_allwvf}
\end{figure}

@@

We perform the same analysis using the design sensitivity curve of AdV and expected sensitivity curves for third-generation 
GW detectors. The results for the former are reported in Table \ref{tab:results} and are discussed below. We focus now on third-generation detectors, presenting our findings in Table \ref{tab:results} and in Fig.~\ref{fig:s20--SFHo_all3G}. In Europe the Einstein Telescope project proposes to host in a 10-km equilateral triangle configuration three low-power, low-frequency, cryogenic interferometers as well as three high-power, high-frequency interferometers. Three sensitivity curves, ET-B, ET-C and ET-D corresponding to different options and stages of the project~\cite{Hild_2011} are considered in our study (cf.~Fig.~\ref{fig:spectrum}). The US based project Cosmic Explorer~\cite{reitze2019cosmic} is expecting to reach its design
sensitivity circa 2040 through two phases labeled CE1 and CE2. Their corresponding sensitivity curves are also shown in Fig.~\ref{fig:spectrum}.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/dist_allwvfs_2G3G}
  \caption{Maximal distance at which the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ is reconstructed
    with good accuracy for all GW detectors analyzed in this study. The values are shown for all eight CCSN waveforms of our sample assuming that the source is optimally oriented with respect to the detector. Some waveform maximal distance markers are not visible because the do overlap (for instance \texttt{s15G} is overlapping with \texttt{s25} for CE2). \tf{Use AdV instead of ADV?}} 
\label{fig:distances}
\end{figure}

>>>>

%\begin{figure}[t]
% \centering
% \includegraphics[width=0.5\textwidth]{plots/spectrum}
% \caption{Amplitude spectral density of the GW detectors aLIGO and AdV at design sensitivity and that of the proposed third-generation detectors Cosmic Explorer and Einstein Telescope. Einstein Telescope sensitivity curve ET-B is obtained pushing second-generation detector technology at its limit. ET-C and ET-D sensitivity curves correspond to a detector configuration where a low-power cryogenic low-frequency interferometer and a high-power room temperature high-frequency interferometer are sharing the same infrastructure~\cite{Hild_2011}. Cosmic Explorer design sensitivity will be achieved in two stages. Stage 1 (CE1) is expected to use the technology developed for the ``A+'' upgrade to aLIGO but scaled up to a 40 km detector while stage 2 (CE2) will implement state-of-the-art technology to decrease quantum and thermal noises~\cite{reitze2019cosmic}.} 
% \label{fig:spectrum}
%\end{figure}

@@

To estimate how accurately we can infer the time evolution of $r(t)$ in the GW data of a single
detector we inject the GW signal for model {\texttt s20S} into 
100 Gaussian noise realisations whose power spectral density (PSD) follows the aLIGO
spectrum~\cite{aLIGOsens:2018}. %The PSD of aLIGO and AdV at design sensitivities, along with those of planned third-generation detectors, are shown in Figure~\ref{fig:spectrum}.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/aLIGO_coverage_allwvfs}
 \caption{Median of the coverage for the eight CCSN waveforms of the {\it test set} embedded in aLIGO noise as a function of the  distance to the source. \mab{The ``no signal'' line shows the median of coverage in absence of any signal. In this example, the median is null and overlaps with the horizontal axis. The blue band boundaries are given by the $\mathrm{5^{th}}$ and $\mathrm{95^{th}}$ percentiles of coverage in absence of any signal.}} \label{fig:aLIGO_cov_allwvf}
\end{figure}

@@

We perform the same analysis using the design sensitivity curve of AdV and expected sensitivity curves for third-generation 
GW detectors. The results for the former are reported in Table \ref{tab:results} and are discussed below. We focus now on third-generation detectors, presenting our findings in Table \ref{tab:results} and in Fig.~\ref{fig:s20--SFHo_all3G}. In Europe the Einstein Telescope project proposes to host in a 10-km equilateral triangle configuration three low-power, low-frequency, cryogenic interferometers as well as three high-power, high-frequency interferometers. Three sensitivity curves, ET-B, ET-C and ET-D corresponding to different options and stages of the project~\cite{Hild_2011} are considered in our study. %(cf.~Fig.~\ref{fig:spectrum}).
The US based project Cosmic Explorer~\cite{reitze2019cosmic} is expecting to reach its design
sensitivity circa 2040 through two phases labeled CE1 and CE2. %Their corresponding sensitivity curves are also shown in Fig.~\ref{fig:spectrum}.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/dist_allwvfs_2G3G}
  \caption{Maximal distance at which the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ is reconstructed
    with good accuracy for all GW detectors analyzed in this study. The values are shown for all eight CCSN waveforms of our sample assuming that the source is optimally oriented with respect to the detector. Some waveform maximal distance markers are not visible because the do overlap (for instance \texttt{s15G} is overlapping with \texttt{s25} for CE2).} 
\label{fig:distances}
\end{figure}
